---
title: "pop-growth-simple-aglB"
output:
  html_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

---
title: "Pop_growth_simple_AV"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# This code takes high throughput microbial growth data from a Bioscreen C instrument, calculates growth rate and stats, and plots the data.

## First upload the data and set up the functions

### Prepare coding environment
```{r}
# fresh start
rm(list = ls())

# load required libraries
library(plyr)
library(tidyr)
library(dplyr)
library(tibble)
library(viridis)
library(ggplot2)
library(tinytex)
library(latexpdf)
library(data.table)
```

### Import the data
```{r}
## read in metadata file. Change directory to data file as needed.
mt<-read.csv('~/data/20190215_aglB_repeat_key_AKS.csv', sep =",", header = TRUE)

mt0 <- mt[-1, -(1:2)]

## read in datafile
dt <- read.csv('~/data/20190215_aglB_repeat.csv', sep = ",", header=FALSE, fileEncoding="UTF-16LE")
        
dt0 <- dt[-(1:2), -2]
colnames(dt0) <- c('time', dt0[1, -1])
dt0 <- dt0[-1, ]
```

### Prepare the data

```{r}
## extract hours, mins, secs from the time column, convert to decimal format
hou <-
  as.numeric(lapply(strsplit(as.character(dt0$time), ':'), '[[', 1))
min <-
  as.numeric(lapply(strsplit(as.character(dt0$time), ':'), '[[', 2))
sec <-
  as.numeric(lapply(strsplit(as.character(dt0$time), ':'), '[[', 3))
## set total time to hours
dt0$time <- (hou) + min / 60 + (sec / 3600)
```

```{r}
# melt the dataframe to long format
m_dt <- melt(dt0, id = 'time')
## bring together the metadata and data files into one data frame
mtdt <- merge(m_dt, mt0, by = 'variable')

```

### Prepare the data for calculating statistics, export to supplementary table.
```{r}
#remove data not relevant to the current study
mtdt <- subset (mtdt, !media == "5M_CM_uracil")

#subtract blanks
#mtdt.strains <- subset(mtdt, strain == "ura3" | strain =="aglB") 
#mtdt.blanks<-subset(mtdt, strain == "blank")
#mtdt.bsub<-mtdt.strains$value - mean(mtdt.blanks$value)
#correct for pathlength difference between bioscreen and a typically cuvette-based spec

#remove blanks
mtdt <- subset (mtdt, !strain == "blank")
mtdt.path<-mtdt$value*2.44

mtdt$value.corr<-mtdt.path

#Make supplementary table S2
write.table (mtdt[ c(-1,-4)], file = "Zaretsky.ST2.txt", sep = "\t", row.names = FALSE, col.names = c("time (hrs)", "OD600", "strain", "medium", "biorep", "techrep", "OD600.corrected"))
#time cutoff for later use in calculating growth rates
t_mtdt <- mtdt[!(mtdt$time <= 10 | mtdt$time >= 30), ]
head(t_mtdt)
```


### Calculate some statistics
```{r}

## summarize all values (get the mean value of growth curve over time for each strain across bioreps, media)
tb_mtdt<- ddply(mtdt, c( 'time',  'strain'),
                plyr::summarise,
                value_mn = mean(value.corr),
                value_sd = sd(value.corr))
head(tb_mtdt)

#summarize values within the 0.5 - 30 hour window, average across techreps but not bioreps
tb_mtdt_reps<- ddply(t_mtdt, c( 'time',  'strain', 'biorep'),
                plyr::summarise,
              value_mn = mean(value.corr),
               value_sd = sd(value.corr))
head(tb_mtdt_reps)
```

### Plot average curves with sd for Figure 4A, Zaretsky et al.
```{r}
ggplot(tb_mtdt, aes(x=time, y=value_mn, group=strain, colour = strain))+
  theme_minimal()+
  labs(x="Time (hours)", y="OD600")+
  scale_y_log10(limits=c(0.2, 3.0))+
  geom_line(stat="smooth",method = "loess",# formula = y~x,
              size = 1 )+
  geom_line(aes(x=time, y = (value_mn-value_sd)),
              stat="smooth",method = "loess",# formula = y~x,
              size = 3,
              alpha = 0.2) +
   geom_line(aes(x=time, y = (value_mn+value_sd)),
              stat="smooth", method = "loess",# formula = y~x,
              size = 3,
              alpha = 0.2) +
  scale_color_manual(values=c(ura3="black",aglB = "blue"))
```

### Calculate growth rates
```{r}

#log transform the summarized data, 10-30 h window
tb_mtdt_reps$log <- log2(tb_mtdt_reps$value_mn)

#make separate data frames for each strain, biorep
ura<-subset (tb_mtdt_reps, tb_mtdt_reps$strain == "ura3")
aglB<-subset (tb_mtdt_reps, tb_mtdt_reps$strain == "aglB" )
ura.a<-subset (ura, biorep == "A")
ura.b<-subset (ura, biorep == "B")
ura.c<-subset (ura, biorep == "C")
ura.d<-subset (ura, biorep == "D")
ura.e<-subset (ura, biorep == "E")

aglB.a<-subset (aglB, biorep == "A")
aglB.b<-subset (aglB, biorep == "B")
aglB.c<-subset (aglB, biorep == "C")
aglB.d<-subset (aglB, biorep == "D")
aglB.e<-subset (aglB, biorep == "E")

#find the slope from the linear model fit to the data for first biorep of ura3 as a test
slope.uraA<-lm(ura.a$log ~ ura.a$time)
summary (slope.uraA)
uraA.coefs<-slope.uraA$coefficients

#see https://feliperego.github.io/blog/2015/10/23/Interpreting-Model-Output-In-R for interpretation of lm summary

#So far, this suggests that a simple linear model fits the exponential phase well: growth rate of 0.08 / h is ~11.8 hr doubling time. Also R2 is 0.99. 
```

### Calculate growth rates using simple linear model for all bioreps, each strain
```{r}
slope.uraB<-lm(ura.b$log ~ ura.b$time)
uraB.coefs<-slope.uraB$coefficients

slope.uraC<-lm(ura.c$log ~ ura.c$time)
uraC.coefs<-slope.uraC$coefficients

slope.uraD<-lm(ura.d$log ~ ura.d$time)
uraD.coefs<-slope.uraD$coefficients

slope.uraE<-lm(ura.e$log ~ ura.e$time)
uraE.coefs<-slope.uraE$coefficients

slope.aglBA<-lm(aglB.a$log ~ aglB.a$time)
aglBA.coefs<-slope.aglBA$coefficients

slope.aglBB<-lm(aglB.b$log ~ aglB.b$time)
aglBB.coefs<-slope.aglBB$coefficients

slope.aglBC<-lm(aglB.c$log ~ aglB.c$time)
aglBC.coefs<-slope.aglBC$coefficients

slope.aglBD<-lm(aglB.d$log ~ aglB.d$time)
aglBD.coefs<-slope.aglBD$coefficients

slope.aglBE<-lm(aglB.e$log ~ aglB.e$time)
aglBE.coefs<-slope.aglBE$coefficients

all.lm.coefs<-as.data.frame(cbind(uraA.coefs,uraB.coefs,uraC.coefs,uraD.coefs,uraE.coefs,aglBA.coefs,aglBB.coefs,aglBC.coefs, aglBD.coefs, aglBE.coefs))
rownames(all.lm.coefs)<-c("Intercept", "slope")

#make a table out of all the linear model coefficient values
all.lm.coefs<-as.data.frame(t(all.lm.coefs))

#set the row names as a column variable
setDT(all.lm.coefs, keep.rownames = "strain.rep")

```

### plot slopes (growth rates)
```{r}
#set up the data frame
all.lm.coefs$strain<- c("ura3", "ura3", "ura3","ura3","ura3", "aglB", "aglB","aglB","aglB","aglB")
mean.rates<-ddply (all.lm.coefs, c('strain'), plyr::summarise, mean.slope = mean(slope), sd.slope = sd(slope) )

# plot the summary stats
ggplot(data = mean.rates, aes(x = strain, y = mean.slope, fill = strain) ) +
  xlab('strain') + 
  ylab('growth rate (per h)')+
  geom_bar(stat='identity', width = 0.5) +
  theme_minimal() +
  geom_errorbar(aes(ymin=mean.slope - sd.slope, ymax=mean.slope + sd.slope), width=0.1,
              position=position_dodge(.9)) +
  scale_fill_manual("strain", values = c(ura3 = "grey", aglB = "blue")) + 
  scale_x_discrete(limits=c("ura3", "aglB"))
```

### Are the differences in growth significant?
```{r}
ura.coefs<-subset (all.lm.coefs, strain == "ura3")
agl.coefs<-subset (all.lm.coefs, strain == "aglB")
t.test (ura.coefs$slope, agl.coefs$slope, paired = FALSE)

#yes, the means of the two samples are significantly different. These p-values are reported in the manuscript text.
```
